---
title: "Rage and the Machines"
subtitle: "A Machine Learning Approach to Analyzing Gun Violence in the US"
author: "Thadryan Sweeney"
date: "April 2, 2018"
output:
  pdf_document: default
  html_document: default
---
# Introduction 

 "year"     
 "month"     
 "intent"    
 "police"   
 "sex"      
 "age"     
 "race"    
 "hispanic" 
 "place"    
 "education"

# Exploring the Dataset

```{R Collecting and Inspecting the Data}

# original 
o.data <- read.csv("rfImputedGunData.csv")

# print out a summary and a list of catagories in each columns 
str(o.data)

# print out names and levels (skip ID)
message("Column names and their levels")
for( i in 2:ncol(o.data)) {
  if((i != 7) & (i != 9)) { # skip lines we don't need
    message(names(o.data[i]))
    print(levels(as.factor(o.data[,i])))
  }
}
```

## Basic Summary of Original Data
```{R Basic summary}

# get a summary of the working data 
summary(o.data)
```

# Managing the Dataset 

Let's make a copy of the original data (in case we want to look back at it) to clean up into our working data. 

We don't need the case ID number "X", it's just the row number, we'll start by getting rid of that. (Classification of hispanic)

```{R Copy and start to prepate the dataset}

# copy to raw data 
data <- o.data

# remove X
data$X <- NULL

# remove seprate classifications of hispanic 
data$hispanic <- NULL
```


# Basic Proportions

## Proportions by Year

Let's get a sense of the makeup of the dataset. We will work our way through the variables

```{R Proportions by Year}

# visualization libraries 
library(ggplot2)
library(reshape2)
library(kableExtra)

# how are deaths distributed over the three years of the dataset?
prop.table(table(data$year))

# chart the deaths by year
ggplot(data, aes(x = factor(1), fill = factor(year)))    + 
                                geom_bar(width = 1)      + 
                                coord_polar(theta = "y") +
                                scale_fill_hue(l = 45, c = 55)
```

The output shows a nearly perfect split between the three years. 
## Proportions by Month
```{R Proportions by Month}

# what proportions of the deaths by month
prop.table(table(data$month))

ggplot(data, aes(month)) + geom_bar(fill = "purple")
```

As with the years, the datset is divded fairly evenly by month save February. While February is the shortest month, it is not apparent if that is enough to explain the discrepancy. We'll talk more about it the appendix. 

####Also I need to fix the x axis on this guy.

## Proportions by Intent
```{R Proportions by Intent}

# what proportions of the deaths are homicide?
prop.table(table(data$intent))

# chart the deaths by intent - barplot
ggplot(data, aes(intent)) + geom_bar(fill = "purple")

# pie chart 
ggplot(data, aes(x = factor(1), fill = factor(intent)))  + 
                                geom_bar(width = 1)      + 
                                coord_polar(theta = "y") +
                                scale_fill_hue(l = 45, c = 55)
```

This is more revealing. The scuicide catagory make up a dominant majority over the other catagories. 

## Proportions by Police

Let's look at police involvment:

```{R Chart of Deaths by Police}

# what are the proportions of police involved shootings
prop.table(table(data$police))

# char the deaths by intent 
ggplot(data, aes(x = factor(1), fill = factor(police)))  + 
                                geom_bar(width = 1)      + 
                                coord_polar(theta = "y") +
                                scale_fill_hue(l = 45, c = 55)
```

Police involved shootings make up just under 2% of the dataset. 

## Proportions by Sex
```{R Proportions by Sex}

# what are the proportions of the dataset by sex
prop.table(table(data$sex))

# chart the deaths by sex - bar plot
ggplot(data, aes(sex)) + geom_bar(fill = "purple") 

# pie chart 
ggplot(data, aes(x = factor(1), fill = factor(sex)))     + 
                                geom_bar(width = 1)      + 
                                coord_polar(theta = "y") +
                                scale_fill_hue(l = 45, c = 55)
```

The majority of the entries in the dataset are of of type "Male".

## Proportions by Age
```{R Proportions by Age}

# chart the deaths by intent 
prop.table(table(data$age))
ggplot(data, aes(age)) + geom_bar(fill = "black") + 
                         scale_x_continuous(breaks = round(seq(min(data$age), max(data$age), by = 5),1)) 
```
```{R Age In Ranges}

# create ranges 
data$age_range[data$age >=  0 & data$age <= 15 ] <- "0-14"
data$age_range[data$age >= 15 & data$age <= 20 ] <- "15-19"
data$age_range[data$age >= 20 & data$age <= 29 ] <- "20-29"
data$age_range[data$age >= 30 & data$age <= 39 ] <- "30-39"
data$age_range[data$age >= 40 & data$age <= 49 ] <- "40-49"
data$age_range[data$age >= 50 & data$age <= 59 ] <- "50-59"
data$age_range[data$age >= 60                  ] <- "60+"

ggplot(data, aes(age_range)) + geom_bar(fill = "purple")
```
47 2012     6      Suicide      0   M 39.00000 White            Home          BA+      <NA>
48 2012     7     Homicide      0   M 59.00000 White            Home       HS/GED      <NA>

## Proportions by Race
```{R Proportions by Race}

# what are the proportions of the dataset by race?
prop.table(table(data$race))

# chart the deaths by race 
ggplot(data, aes(x = factor(1), fill = factor(race)))    + 
                                geom_bar(width = 1)      + 
                                coord_polar(theta = "y") +
                                scale_fill_hue(l = 45, c = 55)
```

## Proportions by Place
```{R Proportions by Place}

# what are the proportions of the dataset by place?
prop.table(table(data$place))

# chart the deaths by race 
ggplot(data, aes(place)) + geom_bar(fill = "navy") + scale_x_discrete(labels = abbreviate)
```

This will be the subject of further investigation, with and without machine approaches, later in the report.

ggplot(data = UFFI,aes(Year_Sold, Sale_Price))+geom_bar(aes(fill=UFFI_contain),stat="identity",position = "dodge")

# Data by the Year

We will not dwell for too long on the year; out initial investigations into the proportions of the dataset suggest that will be pretty consistent. Still, in the name of due dilligence and potentially identifiying emerging trends, we will look at some summary tables and charts.  

```{R More Trends by Year}

# get dataframes by year
data12 <- data[which(data$year == "2012"), ]
data13 <- data[which(data$year == "2013"), ]
data14 <- data[which(data$year == "2014"), ]
```

## Month by Year

Let's look at a simple numeric table of deaths from year to year in the dataset
```{R Month}

# 2012 months
summary(as.factor(data12$month))
# 2013 months
summary(as.factor(data13$month))
# 2014 months 
summary(as.factor(data14$month))
```

The output suggests that there isn't too much variations in the months from year to year. Let's make some visuals:

```{R Months by Year}

# extract month data 
d12 <- data.frame(summary(as.factor(data12$month)))
d13 <- data.frame(summary(as.factor(data13$month)))
d14 <- data.frame(summary(as.factor(data14$month)))

# set months
month <- c(1,2,3,4,5,6,7,8,9,10,11,12)
month.data <- cbind(month,d12,d13,d14)
colnames(month.data) <- c("month", "2012","2013","2014")

# make a new dataframe 
month.data <- melt(month.data, id.vars = "month")

# plot the results on a line graph
ggplot(month.data, aes(month,value, col =  variable)) + 
  geom_line() +
  # set x and y limits 
  scale_y_continuous(limits = c(2250,3250), breaks = seq(2250, 3250, by = 250)) +
  scale_x_continuous(breaks = seq(1,12, by = 1))
```

There appears to be a drop in the deaths during February. Some investigation suggest that it being the shortest month doesn't entirely explain this. We will talk about this in the appendix. 

## Define Chart Function to Format Numeric data 

To save time we will write a generic chart function to make kable charts from data in the the data12, data13, data14 dataframes.

```{R Generic Cable Chart Function}

# define a chart data 
chart_data <- function(catagory) 
{
  # collect education summaries 
  d12 <- as.data.frame(summary((data12[,catagory])))
  d13 <- as.data.frame(summary((data13[,catagory])))
  d14 <- as.data.frame(summary((data14[,catagory])))

  # frame and name education data 
  chart.data <- cbind(d12, d13, d14)

  colnames(chart.data) <- c("2012","2013","2014")

  # chart education data 
  kable(chart.data) %>% 
    kable_styling(position = "center", full_width = TRUE) %>% 
    row_spec(0, bold = TRUE)
}
```


## Intent by Year
```{R Intent}

chart_data("intent")
```

Shootings involving police are tagged as "1", those no are "0".


```{R Police}

# the function will take them as a factor 
data12$police <- as.factor(data12$police)
data13$police <- as.factor(data13$police)
data14$police <- as.factor(data14$police)

# chart the data 
chart_data("police")
```

There is a slight increase in suicides and a slight decrease in homicides. 

## Race by Year
```{R Race by Year}

# chart race
chart_data("race")
```

Guns deaths of the catagory "Black" decrease slightly while the others seem to remain steady. There is not enough information to know if this is normal variation or part of a larger trend. 

## Sex by Year
```{R Sex by Year}

chart_data("sex")
```

There was a slight increase in female deaths in this time period, though there is not enough information to know if this is normal variation or part of a larger trend. 

## Age by Year
```{R Age by Year}

# 2012 ages
#summary(data12$age)
# 2013 age 
#summary(data13$age)
# 2014 ages 
#summary(data14$age)

#data12$age_range <- as.factor(data12$age_range)
#data13$age_range <- as.factor(data13$age_range)
#data14$age_range <- as.factor(data14$age_range)



# chart_data("age_range") # what the actual fuck 
```

## Place by Year
```{R Place by Year}

# call our function 
chart_data("place")
```
## Education by Year
```{R - Education by Year}

# call function on education
chart_data("education")
```

No obvious trends manifest in the education level of the deaths. What are the the porportions of HS/GED in the general populatitiom?


















# Scrap Workspace


ggplot(d, aes(month,d12, color = "blue")) + 
  # add lines 
  geom_line() + 
  # add labels
  scale_x_discrete(labels = abbreviate) +
  # add 2013 and 2014 
  geom_line(aes(month,d13, color = "green")) +
  geom_line(aes(month,d14, color = "orange")) +
  # add colors and labels 
  scale_colour_manual(values=c("blue"="blue","green"="green","orange"="orange"), 
    labels = c("2012","2013","2014"),
    breaks=c("blue","green","orange")) +
  # set axis scales 
  scale_y_continuous(limits = c(2000, 3500))
```{r}

library(ggplot2)
library(reshape2)

# set months
mon <- c(1,2,3,4,5,6,7,8,9,10,11,12)
sd <- cbind(mon,d12,d13,d14)
colnames(sd) <- c("month", "2012","2013","2014")

# make a new dataframe 
d <- melt(sd, id.vars = "month")

# plot the results 
ggplot(d, aes(month,value, col =  variable)) + 
  geom_line()                                +
  scale_y_continuous(limits = c(2250,3250), breaks = seq(2250, 3250, by = 250)) +
  scale_x_continuous(breaks = seq(1,12, by = 1))
```

## proportion_of function
```{R proportion_of() Function}

# allows for string interpolation 
library(stringr)

# define th proportion of to act on df column and target to count 
proportion_of <- function(df, catagory, target, message = TRUE)
{
  # if message option is selected 
  if(message == TRUE) {
  # print a message for clarity 
  message(str_interp("Percent of column '${catagory}' for '${target}'"))
  }
  # return the proportion of the target in the column of the data
  return(  length(which(df[catagory] == target))/nrow(df)  *100)
}

# test the function
#print(proportion_of(data, "police", "1"))
#print(proportion_of(data, "race", "Black"))


```


## Deaths besides suicides 
```{R}


ns.data <- data[which(data$intent != "Suicide"), ]
p.data  <- data[which(data$police == 1), ]

x <- proportion_of(data, "race", "Black")
y <- proportion_of(p.data, "race", "Black")

ns.data <- data[which(data$intent != "Suicide"), ]
p.data  <- data[which(ns.data$police == 1), ]

pie(prop.table(table(ns.data$race)))
```

```{R}

# 
ns.data <- data[which(data$intent != "Suicide"), ]

# the dataset at large is simply "data"

# data where police are involved 
p.data  <- data[which(data$police == 1), ]

# pie chart of racial composition
#message("Dataset in general")
pie(prop.table(table(data$race)))

# pie chart of police data
#message("Police Dataset")
pie(prop.table(table(p.data$race)))

# gun deaths with suicides removed
pie(prop.table(table(ns.data$race)))
```

## this is the racial makeup I was given 
Asian/Pacific Islander 5.25%
White 65.39%
Hispanic 14.79%
Native American/Native Alaskan 1.24%
Black 13.34%


```{r}

#message("Data removing suicide")
#prop.table(table(data$race))
#pie(prop.table(table(data$race)))


message("Data removing suicide")
#summary(ns.data$race)
prop.table(table(ns.data$race))


message("Data involving police")
#summary(p.data$race)
prop.table(table(p.data$race))
```


# What the fuck is up with february
jan:31
feb:28
mar:31

apr:30
may:31
jun:30

jul:31
aug:31
sep:30

oct:31
nov:30
dec:31

2012 was leap year 


```{R WTF is up with February}

d12 <- data.frame(summary(as.factor(data12$month)))
d13 <- data.frame(summary(as.factor(data13$month)))
d14 <- data.frame(summary(as.factor(data14$month)))

months <- c(1,2,3,4,5,6,7,8,9,10,11,12)
days   <- c(31,28,31,30,31,30,31,31,30,31,30,31)
dFebInv <- cbind(months, days, d12, d13, d14)

colnames(dFebInv) <- c("month","days","d12","d13","d14")

for(i in 1:ncol(dFebInv)) {
  dFebInv$expect[i]  <- dFebInv$days[i] * 91.95342
  dFebInv$dif.exp[i] <- (sum(dFebInv$d12[i] + dFebInv$d13[i] + dFebInv$d14[i])/3) - dFebInv$expect[i]
}
dFebInv$z <- scale(dFebInv$expect)
head(dFebInv, n=12)
```




# create a plot starting with 2012
ggplot(d, aes(month,d12, color = "blue")) + 
  # add lines 
  geom_line() + 
  # add labels
  scale_x_discrete(labels = abbreviate) +
  # add 2013 and 2014 
  geom_line(aes(month,d13, color = "green")) +
  geom_line(aes(month,d14, color = "orange")) +
  # add colors and labels 
  scale_colour_manual(values=c("blue"="blue","green"="green","orange"="orange"), 
    labels = c("2012","2013","2014"),
    breaks=c("blue","green","orange")) +
  # set axis scales 
  scale_y_continuous(limits = c(2000, 3500))


(mean(sum(d$d12), sum(d$d13), sum(d$d14))/365) * (28)


```{R}
months <- c(1,2,3,4,5,6,7,8,9,10,11,12)
d <- cbind(months, d12, d13, d14)

colnames(d) <- c("month", "d12","d13","d14")

# create a plot starting with 2012
ggplot(d, aes(month,d12, color = "blue")) + 
  # add lines 
  geom_line() + 
  # add labels
  scale_x_discrete(labels = abbreviate) +
  # add 2013 and 2014 
  geom_line(aes(month,d13, color = "green")) +
  geom_line(aes(month,d14, color = "orange")) +
  # add colors and labels 
  scale_colour_manual(values=c("blue"="blue","green"="green","orange"="orange"), 
    labels = c("2012","2013","2014"),
    breaks=c("blue","green","orange")) +
  # set axis scales 
  scale_y_continuous(limits = c(2000, 3500))


#mean(sum(d$d12), sum(d$d13), sum(d$d14))/365
```