---
title: "scapWork"
author: "Thadryan Sweeney"
date: "May 10, 2018"
output: html_document
---


[ This isn't doing what it is supposed to do yet]
```{R Homicides}

# frame the data by year 
data12 <- na.omit(d[which(d$year == "2012"), ])
data13 <- na.omit(d[which(d$year == "2013"), ])
data14 <- na.omit(d[which(d$year == "2014"), ])


# extract month data 
d12 <- data.frame(summary(as.factor(data12$intent)))
d13 <- data.frame(summary(as.factor(data13$intent)))
d14 <- data.frame(summary(as.factor(data14$intent)))


# make a new dataframe of deaths per month
intent.data <- cbind(d12,d13,d14)

# set new names 
colnames(intent.data) <- c("2012","2013","2014")

# inspect the deaths/month data
kable(intent.data) %>% 
  kable_styling(position = "center", full_width = TRUE) %>% 
  row_spec(0, bold = TRUE)
```

```{R Homicides non.feb}

# frame the data by year 
non.feb.data12 <- na.omit(d[which(d$year == "2012"), ])
non.feb.data13 <- na.omit(d[which(d$year == "2013"), ])
non.feb.data14 <- na.omit(d[which(d$year == "2014"), ])


# extract month data 
d12 <- data.frame(summary(as.factor(non.feb.data12$intent)))
d13 <- data.frame(summary(as.factor(non.feb.data13$intent)))
d14 <- data.frame(summary(as.factor(non.feb.data14$intent)))

# set months
#month <- c(1,2,3,4,5,6,7,8,9,10,11,12)

# make a new dataframe of deaths per month
non.feb.intent.data <- cbind(d12,d13,d14)

# set new names 
colnames(non.feb.intent.data) <- c("2012","2013","2014")

# inspect the deaths/month data
kable(non.feb.intent.data) %>% 
  kable_styling(position = "center", full_width = TRUE) %>% 
  row_spec(0, bold = TRUE)
```

```{R Intent Z-scores}

# make colums for the z-score that start at zero
intent.data$z12 <- 0
intent.data$z13 <- 0
intent.data$z14 <- 0

# iterate through the dataframe 
for(i in 1:nrow(intent.data)) {
  # put the scaled values in the new columns 
  intent.data[i,4:6] <- scale(as.numeric(intent.data[i,1:3]))
}

# label the new data 
colnames(intent.data) <-c("d12", "d13","d14","z12", "z13", "z14")
intent.data <- intent.data[, c("d12","z12","d13","z13","d14", "z14")]

# inspect the deaths/month data
kable(intent.data) %>% 
  kable_styling(position = "center", full_width = TRUE) %>% 
  row_spec(0, bold = TRUE)
```

```{R Scaled Data, eval=FALSE, include=FALSE}

d.intent <- data.frame(d.intent)

# results will be the days-adjusted, scaled value
d.intent$homicide.month.adj <- 0

# make scaled value column set to 0
d.intent$homicide.scaled <- 0 


# divide the results by days in the month to adjust differing lengths
for(i in 1:nrow(d.intent)) {
  # use days vector from previous section
  d.intent$homicide.month.adj[i] <- d.intent$Homicide[i] / days[i]
}

# scale the summary of intent data 
d.intent$homicide.scaled <- scale(d.intent$homicide.month.adj)

# chart the dataframe
kable(d.intent) %>%
  row_spec(2, bold = TRUE, color = "blue")
```


# get deaths by year 
deaths12 <- length(which(c.data$year == 2012))
deaths13 <- length(which(c.data$year == 2013))
deaths14 <- length(which(c.data$year == 2014))

# get lists for the axes
years = c(2012,2013,2014)
year_counts <- c(deaths12, deaths13, deaths14)

# plot the data 
plot(years, year_counts, main = "Total Deaths by Year",ylim = c(32000,32750), xlab = "Year", axes = FALSE)

# modify axes
axis(side = 1, at = years)
axis(side = 2, at = c(seq(32000, 32750, by = 250)))


#for (i in 1:ncol(d)) {
#  d[,i] <- as.numeric(d[,i])
#}
#d$police <- as.factor(d$police)
#df[,c(1,2,5)]
#partitionIndex = createDataPartition(d$d.police, p = 0.7, list = FALSE)

#train <- d[ partitionIndex, ]
#test  <- d[-partitionIndex, ]


#colnames(adult_data)[colSums(is.na(adult_data)) > 0]





# allows for string interpolation 
library(stringr)
# total_message <- str_interp("Total XP: ${total}")
# as
precent_which <- function(df, catagory, target)
{
  message(str_interp("Percent of column '${catagory}' for '${target}'"))
  results <- length(which(df[catagory] == target))/nrow(df)
  return(results*100)
}

precent_which(df = c.data, catagory = "race", target = "Black")

We will now define a function that querys this set of dataframes to give us a side by side comparison.

yby <- function(target)
{
  message(2012)
  print(summary(c.data12[target]))
  message(2012)
  print(summary(c.data13[target]))
  message(2012)
  print(summary(c.data14[target]))
}
```{r}

# define a function to compare elements from the year DFs
year_sbs <- function(target)
{

  # rotate the summmaries in to columns 
  col1 <- transform(as.vector(summary(c.data12[,target])))
  col2 <- transform(as.vector(summary(c.data13[,target])))
  col3 <- transform(as.vector(summary(c.data14[,target])))
  
  # bind the columsm 
  new.df <-cbind(col1, col2, col3)
  
  # rename them
 # colnames(new.df) <- c("2012", "2013", "2014")
  
  # name the rows 
  #print((factor(c.data12[target])))
  
  # display
  return(new.df)
}
```

### Appendix Whatever: validate te proportion_of
```{R Validate the "proportion_of" function - summary, eval=FALSE, include=FALSE}

# get a summary of the racial makeup 
summary(c.data$race)
```
 ## Define a proportion_of() function

We'll be asking a lot of questions about proportions next, so we will define a function that returns the percent of a dataframe in which a given catagory equals a given target, ie, in "this dataframe", what percect of "this colum" is "this trait". By default it will include a message to make it clear what is being analyzed. The validation of this function can be found in appendix *


```{R proportion_of() Function, eval=FALSE, include=FALSE}

# allows for string interpolation 
library(stringr)

# define th proportion of to act on df column and target to count 
proportion_of <- function(df, catagory, target, message = TRUE)
{
  # if message option is selected 
  if(message == TRUE) {
  # print a message for clarity 
  message(str_interp("Percent of column '${catagory}' for '${target}'"))
  }
  # return the proportion of the target in the column of the data
  return(length(which(df[catagory] == target))/nrow(df))
}

# test the function
proportion_of(c.data, "race", "Asian/Pacific Islander", message = FALSE)
```
 R's summary has stated that Asian/Pacific Islander catagory has 1251 entries vs 97218. We will make sure we get the same result before moving on. 
 
```{R Validate the "proportion_of" function - comparison, eval=FALSE, include=FALSE}

# does the result of our function match a manual check
proportion_of(c.data, "race", "Asian/Pacific Islander", message = FALSE) == 1251/97218
```




# get lists for the axes
years = c(2012,2013,2014)
year_counts <- c(deaths12, deaths13, deaths14)

# plot the data 
plot(years, year_counts, main = "Total Deaths by Year",ylim = c(32000,32750), xlab = "Year", axes = FALSE)

# modify axes
axis(side = 1, at = years)
axis(side = 2, at = c(seq(32000, 32750, by = 250)))


```{R Records of unknown intent, eval=FALSE, include=FALSE}

# get a prop table on intent 
prop.table(table(c.data$intent))
```

They make up a very small proportion of out dataset. We will isolate and remove them, updating the clean dataframe.
```{R Remove records of unknown intent, eval=FALSE, include=FALSE}

# isolate undetermined cases 
undetermined <- which(c.data$intent == "Undetermined")

# we will update the dataframe 
c.data <- c.data[-undetermined, ]

# remove the catagory "Undetermined"
c.data$intent <- factor(c.data$intent)
```

We now have a set where all the intents are known and no values are missing.
```{r eval=FALSE, include=FALSE}

#plot(years, intent[1,], ylim = c(0.75*max(intent[1,]), 
 #          1.25*max(intent[1,])), ylab = ("Accidental Deaths"), xlab = ("Year"))

plot(years, intent[1,], ylim = c(0.75*max(intent[1,]),
           1.25*max(intent[3,])), ylab = ("Accidental Deaths"), xlab = ("Year"), type = "line", col = "green")

lines(years, intent[2,], col = "red")
lines(years, intent[3,], col = "blue")

```




```{r eval=FALSE, include=FALSE}

library(ggplot2)
library(reshape2)
#dfm <- melt(intent)
df <- melt(intent)  #the function melt reshapes it from wide to long
df$rowid <- colnames(intent) #add a rowid identifying variable
ggplot(df, aes(variable, value, group=factor(rowid))) + geom_line(aes(color=factor(rowid)))
```


```{r eval=FALSE, include=FALSE}

by_year = as.data.frame(transform(years))
by_year = cbind(byd,year_counts)
colnames(by_year) <- c("Year", "Deaths")

other <- data.frame(Year = c(2012, 2013, 2014) , Deaths= c(32150, 32175, 32330))


ggplot(data = by_year, aes(y = Deaths, x = Year)) + geom_line(color = "blue")+ ylim(32100,32750) +
  geom_line(data = other, color = "red")

#+ ylim(32100,32750)
#  geom_line(data = other, color = "red")
  
```


# Enter the Machines 

Fortunately and unfortunately, our dataset it very large, so we will experiment with a subset of it. We'll shoot for around 1/5th of it. Keep in mind though, that's still 20k. For this we will import R's classic caret library and see how we do with a few machines. 


### TO DO
The Allergory of the police data
```{r}

y.police <- c.data[which(c.data$police == 1), ]
n.police <- c.data[which(c.data$police == 0), ]

length(which(is.na(y.police)))
length(which(is.na(n.police)))

```

```{R Remove missing values}

library(missForest)

# make a copy of the data 
c.data <- o.data

#
mf.data <- missForest(c.data, maxiter = 5, verbose = TRUE)
summary(mf.data)
c.data <- data.frame(mf.data$ximp)
```



